---

title: "Kallisto RNA-seq Analysis"
author: "Mohamed Ashraf Ahmed Ali"
date: "`r Sys.Date()`"
output:
html_document:
toc: true
toc_float: true
number_sections: true
code_folding: show
theme: readable
---------------

# Introduction

This R Markdown document presents the analysis of RNA-seq data generated from a Kallisto-based transcript quantification workflow. Kallisto abundance files were imported into R using TxImport, followed by transcript-to-gene summarization using human gene annotation.

The analysis includes data preprocessing, filtering and TMM normalization, exploratory data analysis, hierarchical clustering, principal component analysis (PCA), differential gene expression analysis using the limma-voom framework, clustering of differentially expressed genes into modules, and functional enrichment analysis using Gene Ontology (GO) enrichment and Gene Set Enrichment Analysis (GSEA).

The workflow was performed using R version 4.6.1 on Windows 11.

---

## R packages used

A variety of R packages was used for this analysis.  All graphics and data wrangling were handled using the [tidyverse suite of packages](https://www.tidyverse.org/).  All packages used are available from the Comprehensive R Archive Network (CRAN), Bioconductor.org, or Github.

---

# Step 1: Importing count data into R

After read mapping with Kallisto, [TxImport](https://bioconductor.org/packages/release/bioc/html/tximport.html) was used to read kallisto outputs into the R environment. Annotation data from the `EnsDb.Hsapiens.v86` package was used to summarize data from transcript-level to gene-level.

```{r setup}
#############################
# R version 4.6.1 (2026-06-24 ucrt)
# Platform: x86_64-w64-mingw32/x64
# Running under: Windows 11 x64 (build 26200)
# 
# Matrix products: default
# LAPACK version 3.12.1
# 
# locale:
# [1] LC_COLLATE=English_United States.utf8  LC_CTYPE=English_United States.utf8   
# [3] LC_MONETARY=English_United States.utf8 LC_NUMERIC=C                          
# [5] LC_TIME=English_United States.utf8   
##############################
```

## Loading packages

```{r load-packages}
# load packages----
library(rhdf5)
library(tidyverse) 
library(tximport) 
library(ensembldb) 
library(EnsDb.Hsapiens.v86) 
library(datapasta)
library(edgeR)
library(matrixStats)
library(cowplot)
library(DT)
library(plotly)
library(gt)
library(htmlwidgets)
library(limma)
library(RColorBrewer)
library(gplots)
library(d3heatmap)
library(GSEABase)
library(Biobase)
library(GSVA)
library(gprofiler2)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
```

## Reading the study design

```{r study-design}
# read in your study design ----
targets <- read_tsv("studydesign.txt")

#creating file paths for the abundance files generated from kallisto
path <- file.path(targets$sample, "abundance.tsv") 

#checking if the paths exists
all(file.exists(path))
```

## Retrieving transcript annotation

```{r transcript-annotation}
#getting annotations using organism-specific packages ----
Tx <- transcripts(EnsDb.Hsapiens.v86, columns = c("tx_id", "gene_name"))
Tx <- as_tibble(Tx)
Tx <- dplyr::rename(Tx, target_id = tx_id)
Tx <- dplyr::select(Tx, target_id, gene_name)
```

## TxImport

```{r step-1-tximport}
# import Kallisto transcript counts into R using Tximport ----
Txi_gene <- tximport(path, 
                     type = "kallisto", 
                     tx2gene = Tx, 
                     txOut = FALSE, 
                     countsFromAbundance = "lengthScaledTPM",
                     ignoreTxVersion = TRUE)
```

---

# Step 2: Preprocessing

The imported expression data were processed to obtain abundance values, estimated counts, and sample labels. Data were then filtered to remove lowly expressed genes and normalized using the TMM method implemented in edgeR.

## Extracting expression data

```{r step2-dataWrangling}
#Extracting abundance, estimated counts, and samples' labels
myTPM <- Txi_gene$abundance
myCounts <- Txi_gene$counts
sampleLabels <- targets$sample

# Summary stats for the data ----
myTPM.Stats <- transform(myTPM,
                         SD=rowSds(myTPM),
                         Ave=rowMeans(myTPM),
                         Med=rowMedians(myTPM)
  
)

#Creating DGElist from myCounts ----
myDGEList <- DGEList(myCounts)
save(myDGEList, file = "myDGEList")

# use the 'cpm' function from EdgeR to get counts per million
cpm <- edgeR::cpm(myDGEList) 
colSums(cpm)
log2.cpm <- edgeR::cpm(myDGEList, log=TRUE)

# 'coerce' data matrix to a dataframe
log2.cpm.df <- as_tibble(log2.cpm, rownames = "geneID")

# add sample names to this dataframe
colnames(log2.cpm.df) <- c("geneID", sampleLabels)
View(log2.cpm.df)
```

## Impact of filtering and normalization

```{r step2-filtering}
#from the messy to tidy data using tidyr package
log2.cpm.df.pivot <- pivot_longer(log2.cpm.df,
                                  cols = HS01:CL13, 
                                  names_to = "samples", 
                                  values_to = "expression")

# now it is easy to plot this pivoted data
log2.cpm.df.pivot <- ggplot(log2.cpm.df.pivot) +
  aes(x=samples, y=expression, fill=samples) +
  geom_violin(trim = FALSE, show.legend = FALSE) +
  stat_summary(fun = "median", 
               geom = "point", 
               shape = 95, 
               size = 10, 
               color = "black", 
               show.legend = FALSE) +
  labs(y="log2 expression", x = "sample",
       title="Log2 Counts per Million (CPM)",
       subtitle="unfiltered, non-normalized",
       caption=paste0("produced on ", Sys.time())) +
  theme_bw()

dir.create("plots", showWarnings = FALSE)
ggsave("plots/log2.cpm.df.pivot.png", plot = log2.cpm.df.pivot, width = 8, height = 6, dpi = 300)
```

### Data filtration

```{r step2-data-filtration}
# Data filtration ----
#first, take a look at how many genes or transcripts have no read counts at all
table(rowSums(myDGEList$counts==0)==10)

keepers <- rowSums(cpm>1)>=5

#use base R's simple subsetting method to filter your DGEList based on the logical produced above
myDGEList.filtered <- myDGEList[keepers,]
dim(myDGEList.filtered)

log2.cpm.filtered <- edgeR::cpm(myDGEList.filtered, log=TRUE)
log2.cpm.filtered.df <- as_tibble(log2.cpm.filtered, rownames = "geneID")
colnames(log2.cpm.filtered.df) <- c("geneID", sampleLabels)

# pivot this FILTERED data, just as you did earlier
log2.cpm.filtered.df.pivot <- pivot_longer(log2.cpm.filtered.df, 
                                           cols = -1, 
                                           names_to = "samples", 
                                           values_to = "expression") 

log2.cpm.filtered.df.pivot <- ggplot(log2.cpm.filtered.df.pivot) +
  aes(x=samples, y=expression, fill=samples) +
  geom_violin(trim = FALSE, show.legend = FALSE) +
  stat_summary(fun = "median", 
               geom = "point", 
               shape = 95, 
               size = 10, 
               color = "black", 
               show.legend = FALSE) +
  labs(y="log2 expression", x = "sample",
       title="Log2 Counts per Million (CPM)",
       subtitle="filtered, non-normalized",
       caption=paste0("produced on ", Sys.time())) +
  theme_bw()

ggsave("plots/log2.cpm.filtered.df.pivot.png", 
       plot = log2.cpm.filtered.df.pivot, width = 8, height = 6, dpi = 300)
```

### Data normalization

```{r step2-normalization}
# Data normalization ----
myDGEList.filtered.norm <- calcNormFactors(myDGEList.filtered, method = "TMM")

# use the 'cpm' function from EdgeR to get counts per million from your normalized data
log2.cpm.filtered.norm <- edgeR::cpm(myDGEList.filtered.norm, log=TRUE)
log2.cpm.filtered.norm.df <- as_tibble(log2.cpm.filtered.norm, rownames = "geneID")
colnames(log2.cpm.filtered.norm.df) <- c("geneID", sampleLabels)
# pivot this NORMALIZED data
log2.cpm.filtered.norm.df.pivot <- pivot_longer(log2.cpm.filtered.norm.df, 
                                                cols = -1, 
                                                names_to = "samples", 
                                                values_to = "expression") 


log2.cpm.filtered.norm.df.pivot <- ggplot(log2.cpm.filtered.norm.df.pivot) +
  aes(x=samples, y=expression, fill=samples) +
  geom_violin(trim = FALSE, show.legend = FALSE) +
  stat_summary(fun = "median", 
               geom = "point", 
               shape = 95, 
               size = 10, 
               color = "black", 
               show.legend = FALSE) +
  labs(y="log2 expression", x = "sample",
       title="Log2 Counts per Million (CPM)",
       subtitle="filtered, TMM normalized",
       caption=paste0("produced on ", Sys.time())) +
  theme_bw()

ggsave("plots/log2.cpm.filtered.norm.df.pivot.png",
       plot = log2.cpm.filtered.norm.df.pivot, width = 8, height = 6, dpi = 300)

#put the three figures together using plot_grid
raw.filtered.norm <- plot_grid(log2.cpm.df.pivot,
                                 log2.cpm.filtered.df.pivot,
                                 log2.cpm.filtered.norm.df.pivot,
                                 labels = c('A', 'B', 'C'),
                                 label_size = 12)

ggsave("plots/raw.filtered.norm.png", 
       plot = raw.filtered.norm,
       width = 8, height = 6, dpi = 300)
```

---

# Step 3: Data exploration

The processed expression data were explored using hierarchical clustering and principal component analysis (PCA) to assess relationships between samples and visualize overall patterns of variation.

## Study groups

```{r step3-study-groups}
# Identify variables of interest in study design file ----
group <- factor (targets$group)
```

## Hierarchical clustering

```{r step3-hierarchical-clustering}
# Hierarchical clustering ---------------
distance <- dist(t(log2.cpm.filtered.norm), method = "maximum") 
clusters <- hclust(distance, method = "average") 
Hierarchical_clustering <- plot(clusters, labels=sampleLabels)

png(
  "plots/Hierarchical_clustering.png",
  width = 2400,
  height = 1800,
  res = 300
)

plot(clusters, labels = sampleLabels)

dev.off()
```

## Principal component analysis

```{r step3-pca}
# Principal component analysis (PCA) -------------
pca.res <- prcomp(t(log2.cpm.filtered.norm), scale.=F, retx=T)
pc.var<-pca.res$sdev^2 
pc.per<-round(pc.var/sum(pc.var)*100, 1)

ls(pca.res)
summary(pca.res) 
pca.res$rotation 
pca.res$x 

scree.df <- data.frame(
  PC = factor(
    paste0("PC", 1:length(pc.var)),
    levels = paste0("PC", 1:length(pc.var))
  ),
  Variance = pc.var,
  Percentage = pc.per
)

Scree.Plot <- ggplot(scree.df[1:10, ], aes(x = PC, y = Variance)) +
  geom_col() +
  labs(
    title = "PCA Scree Plot",
    x = "Principal Component",
    y = "Variance"
  ) +
  theme_bw()

Scree.Plot

ggsave("plots/Scree.Plot.png",
       plot = Scree.Plot,
       width = 8, height = 6, dpi = 300)
```

## PCA visualization

```{r step3-pca-visualization}
# PCA Visualization -----
pca.res.df <- as_tibble(pca.res$x)
PCA.Plot <- ggplot(pca.res.df) +
  aes(x=PC1, y=PC2, label=sampleLabels, colour = group) +
  geom_point(size=4) +
  # geom_label() +
  # stat_ellipse() +
  xlab(paste0("PC1 (",pc.per[1],"%",")")) + 
  ylab(paste0("PC2 (",pc.per[2],"%",")")) +
  labs(title="PCA plot",
       caption=paste0("produced on ", Sys.time())) +
  # coord_fixed() +
  theme_bw()

ggsave("plots/PCA.Plot.png",
       plot = PCA.Plot,
       width = 8, height = 6, dpi = 300)
```

## PCA small multiples

```{r step3-pca-small-multiples}
# Create a PCA 'small multiples' chart ----
pca.res.df <- pca.res$x[,1:4] %>% 
  as_tibble() %>%
  add_column(sample = sampleLabels,
             group = group)

pca.pivot <- pivot_longer(pca.res.df, 
                          cols = PC1:PC4, 
                          names_to = "PC", 
                          values_to = "loadings") 

PCA.sm.multi <- ggplot(pca.pivot) +
  aes(x=sample, y=loadings, fill=group) + 
  geom_bar(stat="identity") +
  facet_wrap(~PC) +
  labs(title="PCA 'small multiples' plot",
       caption=paste0("produced on ", Sys.time())) +
  theme_bw() +
  coord_flip()

ggsave("plots/PCA.sm.multi.png",
       plot = PCA.sm.multi,
       width = 8, height = 6, dpi = 300)
```

## Data manipulation and exploratory tables

```{r step3-data-manipulation}
# Data manipulation using dplyr package ----
mydata.df <- log2.cpm.filtered.norm.df %>% 
  mutate(healthy.AVG = (HS01 + HS02 + HS03 + HS04 + HS05)/5,
         disease.AVG = (CL08 + CL10 + CL11 + CL12 + CL13)/5,
         #now make columns comparing each of the averages above that you're interested in
         LogFC = (disease.AVG - healthy.AVG)) %>% 
  mutate_if(is.numeric, round, 2)


# Use dplyr 'arrange' and 'select' to sort dataframe based on any variable
mydata.sort <- mydata.df %>%
  dplyr::arrange(desc(LogFC)) %>% 
  dplyr::select(geneID, LogFC)

# Use dplyr "filter" and "select" functions to pick out genes of interest 
mydata.filter <- mydata.df %>%
  dplyr::filter(geneID=="MMP1" | geneID=="GZMB" | geneID=="IL1B" | geneID=="GNLY" | geneID=="IFNG"
                | geneID=="CCL4" | geneID=="PRF1" | geneID=="APOBEC3A" | geneID=="UNC13A" ) %>%
  dplyr::select(geneID, healthy.AVG, disease.AVG, LogFC) %>%
  dplyr::arrange(desc(LogFC))

# filter based on any regular expression
mydata.grep <- mydata.df %>%
  dplyr::filter(grepl('CXCL|IFI', geneID)) %>%
  dplyr::select(geneID, healthy.AVG, disease.AVG, LogFC) %>%
  dplyr::arrange(desc(geneID))
```

## Publication-quality tables

```{r step3-gt-table}
# Produce publication-quality tables using the gt package ----
table1 <- mydata.filter %>%
  gt() %>%
  fmt_number(columns=2:4, decimals = 1) %>%
  tab_header(title = md("**Regulators of skin pathogenesis**"),
             subtitle = md("*during cutaneous leishmaniasis*")) %>%
  tab_footnote(
    footnote = "Deletion or blockade ameliorates disease in mice",
    locations = cells_body(
      columns = geneID,
      rows = c(6, 7))) %>% 
  tab_footnote(
    footnote = "Associated with treatment failure in multiple studies",
    locations = cells_body(
      columns = geneID,
      rows = c(2:8))) %>%
  tab_footnote(
    footnote = "Implicated in parasite control",
    locations = cells_body(
      columns = geneID,
      rows = c(2))) %>%
  tab_source_note(
    source_note = md("Reference: Amorim *et al*., (2019). DOI: 10.1126/scitranslmed.aar3619"))

gtsave(table1, "plots/Tables/table1.png")
```

## Interactive tables

```{r step3-interactive-table}
# Interactive tables using DT package ----
int.table1 <- datatable(mydata.df[,c(1,12:14)], 
          extensions = c('KeyTable', "FixedHeader"), 
          filter = 'top',
          options = list(keys = TRUE, 
                         searchHighlight = TRUE, 
                         pageLength = 10, 
                         lengthMenu = c("10", "25", "50", "100")))

saveWidget(
  int.table1,
  "plots/Tables/int.table1.html",
  selfcontained = TRUE
)
```

## Interactive scatter plot

```{r step3-interactive-scatter}
# Make an interactive scatter plot with plotly -----
# plotting using ggplot
myplot <- ggplot(mydata.df) +
  aes(x=healthy.AVG, y=disease.AVG, 
      text = paste("Symbol:", geneID)) +
  geom_point(shape=16, size=1) +
  ggtitle("disease vs. healthy") +
  theme_bw()

#converting it into interactive plot using plotly
int.plot1 <- ggplotly(myplot)

saveWidget(
  int.plot1,
  "plots/int.diseaseVShealthy_plot.html",
  selfcontained = TRUE
)
```

---

# Step 4: Differential Gene Expression

Differential gene expression analysis was performed using the limma-voom workflow. A design matrix was constructed to represent the healthy and disease groups, followed by modeling of the mean-variance relationship, linear modeling, contrast analysis, and empirical Bayes moderation.

## Design matrix and VOOM transformation

```{r step4-voom}
# Set up design matrix ----
group <- factor(targets$group)
design <- model.matrix(~0 + group)
colnames(design) <- levels(group)

# Model mean-variance trend and fit linear model to data ----
# Use VOOM function from Limma package to model the mean-variance relationship
png(
  "plots/voom_plot.png",
  width = 2400,
  height = 1800,
  res = 300
)

v.DEGList.filtered.norm <- voom(myDGEList.filtered.norm, design, plot = TRUE)

dev.off()
# fit a linear model to your data
fit <- lmFit(v.DEGList.filtered.norm, design)
```

## Contrast and empirical Bayes analysis

```{r step4-contrast}
# Contrast matrix ----
contrast.matrix <- makeContrasts(infection = disease - healthy,
                                 levels=design)

# extract the linear model fit -----
fits <- contrasts.fit(fit, contrast.matrix)

#get bayesian stats for your linear model fit
ebFit <- eBayes(fits)
write.fit(ebFit, file="results/lmfit_results.txt")
```

## Differential expression results

```{r step4-deg-results}
# TopTable to view DEGs -----
# Getting and saving complete results 
dir.create("results", showWarnings = FALSE)

DEG_results <- topTable(
  ebFit,
  adjust = "BH",
  coef = 1,
  number = Inf,
  sort.by = "logFC"
)%>%
  as_tibble(rownames = "geneID")

write.csv(
  DEG_results,
  "results/DEG_results.csv",
  row.names = TRUE
)

# Extracting and saving myTopHits 
myTopHits <- topTable(ebFit, adjust ="BH", coef=1, number=10, sort.by="logFC")

# convert to a tibble
myTopHits.df <- myTopHits %>%
  as_tibble(rownames = "geneID")

gt(myTopHits.df)

write.csv(
  myTopHits.df,
  "results/myTopHits.df.csv",
  row.names = TRUE
)
```

## Volcano plot

```{r step4-volcano}
# Volcano Plots ----
vplot <- ggplot(DEG_results) +
  aes(y=-log10(adj.P.Val), x=logFC, text = paste("Symbol:", geneID)) +
  geom_point(size=2) +
  geom_hline(yintercept = -log10(0.01), linetype="longdash", colour="grey", linewidth=1) +
  geom_vline(xintercept = 1, linetype="longdash", colour="#BE684D", linewidth=1) +
  geom_vline(xintercept = -1, linetype="longdash", colour="#2C467A", linewidth=1) +
  #annotate("rect", xmin = 1, xmax = 12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#BE684D") +
  #annotate("rect", xmin = -1, xmax = -12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#2C467A") +
  labs(title="Volcano plot",
       subtitle = "Cutaneous leishmaniasis",
       caption=paste0("produced on ", Sys.time())) +
  theme_bw()

ggsave("plots/V.plot.png",
       vplot, width = 8, height = 6, dpi = 300)
```

## Interactive volcano plot

```{r step4-interactive-volcano}
# Interactive Volcano plot
int.v.plot <- ggplotly(vplot)

saveWidget(
  int.v.plot,
  "plots/int.v.plot.html",
  selfcontained = TRUE
)
```

## Identifying differentially expressed genes

```{r step4-decideTests}
# decideTests to pull out the DEGs and make Venn Diagram ----
results <- decideTests(ebFit, method="global", adjust.method="BH", p.value=0.01, lfc=2)

# take a look at what the results of decideTests looks like
head(results)
summary(results)
vennDiagram(results, include="up")
```

## Extracting DEG expression data

```{r step4-extract-degs}
# retrieve expression data for your DEGs ----
head(v.DEGList.filtered.norm$E)
colnames(v.DEGList.filtered.norm$E) <- sampleLabels

diffGenes <- v.DEGList.filtered.norm$E[results[,1] !=0,]
head(diffGenes)
dim(diffGenes)
#convert your DEGs to a dataframe using as_tibble
diffGenes.df <- as_tibble(diffGenes, rownames = "geneID")
```

## Interactive DEG table

```{r step4-deg-table}
# create interactive tables to display your DEGs ----
datatable(diffGenes.df,
          extensions = c('KeyTable', "FixedHeader"),
          caption = 'Table 1: DEGs in cutaneous leishmaniasis',
          options = list(keys = TRUE, searchHighlight = TRUE, pageLength = 10, lengthMenu = c("10", "25", "50", "100"))) %>%
  formatRound(columns=c(2:11), digits=2)

#write your DEGs to a file
write_tsv(diffGenes.df,"DiffGenes.txt")
write.csv(diffGenes.df,"results/DiffGenes.csv")
```

---

# Step 5: Module Identification

Differentially expressed genes were further examined using hierarchical clustering to identify groups of genes with similar expression patterns across the samples.

## Clustering DEGs

```{r step5-clustering}
# Cluster DEGs ----
# Clustering the genes (rows)
clustRows <- hclust(as.dist(1-cor(t(diffGenes), method="pearson")), method="complete") 

# Clustering the samples (columns)
clustColumns <- hclust(as.dist(1-cor(diffGenes, method="spearman")), method="complete") 

#Cut the resulting tree and create color vector for clusters
module.assign <- cutree(clustRows, k=2)
module.color <- rainbow(length(unique(module.assign)), start=0.1, end=0.9) 
module.color <- module.color[as.vector(module.assign)] 
```

## DEG heatmap

```{r step5-heatmap}
# Produce a static heatmap of DEGs -------
# Choose color palette
myheatcolors1 <- bluered(75) 
display.brewer.all()
display.brewer.all(colorblindFriendly = TRUE)

#plot the hclust results as a heatmap
png("DEGs_heatmap.png",
    width = 3000,
    height = 2400,
    res = 300)

heatmap.2(diffGenes,
          Rowv = as.dendrogram(clustRows),
          Colv = as.dendrogram(clustColumns),
          RowSideColors = module.color,
          col = rev(myheatcolors1),
          scale = "row",
          labRow = NA,
          density.info = "none",
          trace = "none",
          cexRow = 1,
          cexCol = 1,
          margins = c(8, 20))

dev.off()
```

## Module assignments

```{r step5-module-assignments}
# View modules of co-regulated genes ----
# view color assignments for the different clusters
names(module.color) <- names(module.assign) 

module.assign.df <- as_tibble(as.list(module.assign))
module.assign.pivot <- pivot_longer(module.assign.df, 
                                    cols = everything(), 
                                    names_to = "geneID", 
                                    values_to = "module")

module.assign.pivot <- module.assign.pivot %>%
  mutate(moduleColor = case_when(
    module == 1 ~ "#FF9900",
    module == 2 ~ "#FF0099"))

modules.count <- ggplot(module.assign.pivot) +
  aes(module) +
  geom_bar(aes(fill=moduleColor)) +
  theme_bw()

ggsave("plots/modules.count.png",
       modules.count, width = 8, height = 6, dpi = 300)
```

## Upregulated module

```{r step5-upregulated-module}
#choose a cluster(s) of interest by selecting the corresponding number based on the previous graph
# The genes go up in the disease (Module 2)
modulePick <- 2 
myModule <- diffGenes[names(module.assign[module.assign %in% modulePick]),] 
hrsub <- hclust(as.dist(1-cor(t(myModule), method="pearson")), method="complete") 

# Create heatmap for chosen sub-cluster.
png("plots/UPgenes.module.png",
    width = 3000,
    height = 2400,
    res = 300)

heatmap.2(
  myModule,
  Rowv = as.dendrogram(hrsub),
  Colv = NA,
  labRow = NA,
  col = rev(myheatcolors1),
  scale = "row",
  density.info = "none",
  trace = "none",
  RowSideColors = module.color[module.assign %in% modulePick],
  margins = c(8, 20)
)

dev.off()
```

## Downregulated module

```{r step5-downregulated-module}
# The genes go down in the disease (Module 1)
modulePick.down <- 1 
myModule.down <- diffGenes[names(module.assign[module.assign %in% modulePick.down]),] 
hrsub.down <- hclust(as.dist(1-cor(t(myModule.down), method="pearson")), method="complete") 

# Create heatmap for chosen sub-cluster.
png("plots/DownGenes.module.png",
    width = 3000,
    height = 2400,
    res = 300)

heatmap.2(
  myModule.down,
  Rowv = as.dendrogram(hrsub.down),
  Colv = NA,
  labRow = NA,
  col = rev(myheatcolors1),
  scale = "row",
  density.info = "none",
  trace = "none",
  RowSideColors = module.color[module.assign %in% modulePick.down],
  margins = c(8, 20)
)

dev.off()
```

## Exporting modules

```{r step5-export-modules}
# Export modules for downstream analysis ----
#prints out genes in the order you see them in the cluster
moduleSymbols <- tibble(geneID = rev(hrsub$labels[hrsub$order]))
moduleData <- diffGenes[moduleSymbols$geneID,]
moduleData.df <- as_tibble(moduleData, rownames = "geneSymbol")
write_csv(moduleData.df,"results/module_upRegulated.csv")
```

---

# Step 6: Functional Enrichment

Functional enrichment analysis was performed to investigate biological processes and molecular signatures associated with the identified gene modules. GO enrichment was performed using gProfiler2, while GSEA was performed using clusterProfiler and MSigDB gene sets.

## GO enrichment of upregulated genes

```{r step6-go-up}
# Carry out GO enrichment using gProfiler2 ----
#Enrichment for myModule: the upregulated genes
# GO enrichment analysis for myModule
gost.up <- gost(
  rownames(myModule),
  organism = "hsapiens",
  correction_method = "fdr"
)

# Create static GO enrichment plot
mygostplot.up <- gostplot(
  gost.up,
  interactive = FALSE,
  capped = TRUE
)

# Display publication-quality plot
publish_gostplot(
  mygostplot.up,
  filename = NULL,
  width = NA,
  height = NA
)

# Save myModule GO plot
png(
  "plots/myModule_GO.png",
  width = 3000,
  height = 2400,
  res = 300
)

publish_gostplot(
  mygostplot.up,
  filename = NULL,
  width = NA,
  height = NA
)

dev.off()
```

## GO enrichment of downregulated genes

```{r step6-go-down}
#Enrichment for myModule.down: the downregulated genes
# GO enrichment analysis for myModule.down
gost.down <- gost(
  rownames(myModule.down),
  organism = "hsapiens",
  correction_method = "fdr"
)

# Create static GO enrichment plot
mygostplot.down <- gostplot(
  gost.down,
  interactive = FALSE,
  capped = TRUE
)

# Display publication-quality plot
publish_gostplot(
  mygostplot.down,
  filename = NULL,
  width = NA,
  height = NA
)

# Save myModule.down GO plot
png(
  "plots/myModule.down_GO.png",
  width = 3000,
  height = 2400,
  res = 300
)

publish_gostplot(
  mygostplot.down,
  filename = NULL,
  width = NA,
  height = NA
)

dev.off()
```

## Gene Set Enrichment Analysis

```{r step6-gsea}
# Perform GSEA using clusterProfiler ----
# Reading msigDB collections into R after direct download from GSEA
C2CP <- read.gmt("c2.all.v7.1.symbols.gmt")
# There is an alternative option: to access it using msigdbr package in R

# Data preparation 
mydata.df.sub <- dplyr::select(mydata.df, geneID, LogFC)
# construct a named vector
mydata.gsea <- mydata.df.sub$LogFC
names(mydata.gsea) <- as.character(mydata.df.sub$geneID)
mydata.gsea <- sort(mydata.gsea, decreasing = TRUE)

# run GSEA using the 'GSEA' function from clusterProfiler
set.seed(123) 
myGSEA.res <- GSEA(mydata.gsea, TERM2GENE=C2CP, verbose=FALSE) 
myGSEA.df <- as_tibble(myGSEA.res@result)
```

## Interactive GSEA results

```{r step6-gsea-table}
# view results as an interactive table
datatable(myGSEA.df,
          extensions = c('KeyTable', "FixedHeader"),
          caption = 'Signatures enriched in leishmaniasis',
          options = list(keys = TRUE, searchHighlight = TRUE, pageLength = 10, lengthMenu = c("10", "25", "50", "100"))) %>%
  formatRound(columns=c(2:10), digits=2)
```

## GSEA enrichment plots

```{r step6-gsea-plots}
# create enrichment plots using the enrichplot package
png(
  "plots/GSEA_selected_pathways.png",
  width = 3000,
  height = 2400,
  res = 300
)

gseaplot2(
  myGSEA.res,
  geneSetID = c(9, 23, 28, 35),
  pvalue_table = FALSE,
  title = myGSEA.res$Description[14]
)

dev.off()
```

## GSEA phenotype classification

```{r step6-gsea-phenotype}
# add a variable to this result that matches enrichment direction with phenotype
myGSEA.df <- myGSEA.df %>%
  mutate(phenotype = case_when(
    NES > 0 ~ "disease",
    NES < 0 ~ "healthy"))
```

## GSEA bubble plot

```{r step6-gsea-bubble}
# create 'bubble plot' to summarize y signatures across x phenotypes
phen.sig.bubble.plot <- ggplot(myGSEA.df[1:20, ], aes(x = phenotype, y = ID)) +
  geom_point(aes(size = setSize, color = NES, alpha = -log10(p.adjust))) +
  scale_color_gradient(low = "blue", high = "red") +
  scale_y_discrete(labels = function(x) stringr::str_trunc(x, width = 40)) +
  theme_bw() +
  theme(
    axis.text.y = element_text(size = 8),
    axis.text.x = element_text(size = 10),
    # Fixed aspect ratio (height:width) forces the plot box to stay narrow
    aspect.ratio = 2.2 
  )

# Save with a compact width so the frame doesn't stretch across the canvas
ggsave(
  "plots/phen.sig.bubble.plot.png",
  phen.sig.bubble.plot,
  width = 7,
  height = 7,
  dpi = 300
)
```

---

# End of Analysis

This document contains the complete Kallisto-to-RNA-seq analysis workflow, including data import, preprocessing, exploratory analysis, differential gene expression analysis, module identification, and functional enrichment analysis.
